---
title: "Detecting disclosure of generative-AI use"
output: rmarkdown::html_vignette
description: >
  How rt_ai_pmc() detects whether an article discloses the use (or non-use) of
  generative AI in preparing the manuscript, why it is gated to 2023 onward, and
  how to summarize it across a corpus.
vignette: >
  %\VignetteIndexEntry{Detecting disclosure of generative-AI use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
```

```{r setup}
library(rtransparent)
```

## What this indicator captures

Since 2023, journals and publishers (ICMJE, COPE, Nature, Science, Elsevier and
others) have asked authors to **disclose any use of generative AI** such as
ChatGPT, GPT-4, Copilot, Gemini or other large language models when preparing a
manuscript. `rt_ai_pmc()` detects whether an article carries such a disclosure.

It is deliberately narrow. It targets AI used to **prepare the manuscript**
(writing, editing, language, figures), and counts both directions of the
disclosure:

* a positive disclosure: *"the authors used ChatGPT to improve the readability of
  the manuscript"*;
* a negative disclosure: *"no generative AI was used in the preparation of this
  work"*;
* a dedicated section: a *"Declaration of generative AI"* heading.

It does **not** count AI that is the article's research *method* (for example a
paper that trains a deep-learning classifier, or one that studies ChatGPT as its
subject). Those mention AI heavily but make no statement about using AI to write
the paper, so they are not disclosures. A negative lookahead also keeps the tool
sense of "large language model" from being read as something the authors edited.

## The 2023 year gate

The practice did not exist before 2023, so evaluating it on older articles would
be meaningless. `rt_ai_pmc()` therefore reads the publication `year` and:

* returns `is_ai_pred = NA` for articles before 2023 (and reports the `year`);
* returns `TRUE` or `FALSE` for 2023 onward.

The bundled example article is from 2020, so it returns `NA`:

```{r}
xml_path <- system.file(
  "extdata", "PMID32171256-PMC7071725.xml", package = "rtransparent"
)
ai <- rt_ai_pmc(xml_path, remove_ns = TRUE)
c(year = ai$year, is_ai_pred = ai$is_ai_pred)
```

The three possible values are easy to act on: `TRUE` (a disclosure was found),
`FALSE` (the article is from 2023 or later but carries no disclosure), and `NA`
(the article predates the practice and was not assessed). `rt_summary()` drops
the `NA`s, so a corpus prevalence is computed only over the articles where the
indicator applies.

## In the all-indicators output

`rt_all_pmc()` includes the indicator, so a single pass over a corpus already
carries `year` and `is_ai_pred` alongside COI, funding, registration, novelty
and replication (data and code sharing come from `rt_data_code_pmc()`):

```{r}
all_indicators <- rt_all_pmc(xml_path, remove_ns = TRUE)
all_indicators[, c("pmid", "year", "is_ai_pred")]
```

## Across a corpus

Because the indicator is so new, its corpus-level story is a **trend**: how fast
disclosure is being adopted from 2023 onward. The bundled simulated corpus
`rt_demo` includes an `is_ai_pred` column (`NA` before 2023) for illustration.

```{r}
data(rt_demo)
ai_by_year <- rt_summary(rt_demo, by = "year", indicators = "is_ai_pred")
# Years before 2023 have no assessable articles (all NA), so no denominator;
# keep only the years where the indicator applies.
ai_by_year <- ai_by_year[ai_by_year$n_articles > 0, ]
knitr::kable(
  ai_by_year[, c("year", "n_articles", "n_detected", "percent")],
  digits = 1,
  col.names = c("Year", "Assessed", "Disclosed", "%")
)
```

Only 2023 onward carries a denominator; earlier years have no assessable
articles and are filtered out above. Plotted, the adoption curve is the point of
the indicator (`rt_plot()` drops the empty earlier years automatically):

```{r, eval = has_ggplot, fig.width = 7, fig.height = 3.5, fig.alt = "Line chart of generative-AI-use disclosure prevalence by year from 2023"}
library(ggplot2)
rt_plot(rt_demo, type = "trend", year = "year", indicators = "is_ai_pred") +
  ggtitle("Disclosure of generative-AI use (simulated corpus)")
```

It also sits naturally next to the other indicators in a single prevalence
chart; the AI bar simply reflects the 2023-onward subset:

```{r, eval = has_ggplot, fig.width = 7, fig.height = 3.5, fig.alt = "Bar chart of all transparency indicators including AI-use disclosure"}
rt_plot(rt_demo) + ggtitle("Transparency indicators, including AI-use disclosure")
```

## Notes on precision

On a manual validation of 1,000 open-access PMC articles (almost all 2024-2026),
the indicator flagged about 16% of articles, and inspected positives were
genuine disclosures. The main thing it intentionally avoids is treating articles
that *discuss* ChatGPT as their topic, or *use* AI as a research method, as
disclosures: those are not statements about how the manuscript was written. As
with every other indicator, `rt_ai_pmc()` returns the matched text (`ai_text`)
so a positive call can always be inspected.

For the other indicators and the corpus-summary tools used here, see the
[Introduction](rtransparent.html) and
[Summarizing transparency across a corpus](transparency-summary.html) articles.
