---
title: "Scope and limitations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scope and limitations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`rtransparent` is a pattern-based detector. It is designed for high precision on
the statements it targets, and its predictions come with the exact text that
triggered them so they can be audited. This vignette describes what each
indicator does and does not capture, so results are interpreted correctly.

## What the indicators mean

| Indicator | Detects | Does **not** mean |
|---|---|---|
| Conflicts of interest | A COI **disclosure is present** (including "the authors declare no competing interests") | That a conflict exists |
| Funding | A statement that funding **was received** | Presence of a funding *section* (a "no funding" section is read as absence) |
| Registration | A protocol/trial **registration identifier or statement** | Ethics/IRB approval numbers |
| Novelty | The article **claims** its own work is novel or first | That the work is objectively novel |
| Replication | A replication or external/independent validation was **performed** | An internal train/test split, or future/recommended validation |
| Data sharing | The authors' **own data are made available** (repository, accession, or in-article) | Data merely reused, cited, or available "upon request" |
| Code sharing | The authors' **own analysis code is shared** | Use of third-party software/tools |
| AI disclosure | A statement **discloses** generative-AI use in manuscript preparation (including "no AI was used") | Use of AI as a research method |

Conflicts of interest and AI disclosure are **disclosure-based**: a statement
addressing the topic counts as present, whether the disclosure is positive or
negative. This mirrors how these are reported and counted in the literature.

## Known limitations

- **Language.** Detection is strongest in English. Conflict-of-interest and
  funding statements are also detected in Spanish, Portuguese, French, German
  and Italian; other indicators and other languages are not yet covered.
- **Data availability "upon request".** Data offered only on request are **not**
  counted as shared, reflecting the modern open-data standard. This is stricter
  than some earlier definitions and will report lower data-sharing prevalence
  than tools that count availability-on-request.
- **Novelty and replication are claim detection.** They identify what authors
  *state*, not whether a study is truly novel or a replication succeeded. The
  replication indicator in particular is precision-limited because validation
  language ("validation cohort", "independent") is heavily overloaded; see the
  replication-enriched benchmark in `inst/benchmark/`.
- **Plain text vs XML.** The plain-text detectors share the same logic as the
  PMC XML detectors but cannot use XML-structural cues (tagged funding groups,
  conflict footnotes, section types), so a few statements detectable in XML are
  not detectable in plain text.
- **Accuracy correction.** `rt_summary()` can correct apparent prevalence using
  bundled sensitivity/specificity estimates (`rt_accuracy`). These derive from
  the validation benchmarks; supply your own via `rt_summary(accuracy = )` when
  you have study-specific estimates. AI disclosure is reported uncorrected (its
  prevalence is too low in unselected literature for a stable estimate).

## Output schema

Every per-article detector returns the prediction columns `is_coi_pred`,
`is_fund_pred`, `is_register_pred`, `is_novelty_pred`, `is_replication_pred`,
`is_open_data`, `is_open_code`, and the year-gated `is_ai_pred` (`NA` before
2023), each paired with the extracted text. `rt_all_pmc()` returns all eight for
one file; `rt_all_pmc_dir()` runs a whole directory.

```{r}
library(rtransparent)

res <- rt_all_pmc("article.xml", remove_ns = TRUE)
res[, c("is_coi_pred", "is_fund_pred", "is_open_data", "is_open_code")]
```

## Linking to FAIR assessment

The data- and code-availability links the detector extracts
(`open_data_links`, `open_code_links`) can be passed to FAIR-assessment tooling
such as [`rfuji`](https://github.com/Ammar-Ali-Khan/rfuji) to score the
findability and accessibility of the shared resources.

```{r}
res <- rt_all_pmc("article.xml", remove_ns = TRUE)
links <- strsplit(res$open_data_links, " ; ")[[1]]
# rfuji::assess_fair(links)
```
